---
title: "K00 OU Tracking Fit Report"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)

ANALYSIS_DIR <- if (
  dir.exists("data") &&
    file.exists("data/ou_tracking_fits_yellow_reconstructed_area2x_nonnegative_trackids_min3.csv")
) {
  "."
} else {
  "analyses/K00_GemcitabineExposure_033023"
}
YELLOW_TRACK_DATASET <- "yellow_reconstructed_area2x_nonnegative_tracks"
YELLOW_TRACKS_RDS <- file.path(
  ANALYSIS_DIR,
  "data/tracking_data_yellow_reconstructed_area2x_nonnegative_trackids_min3.rds"
)
YELLOW_OU_FITS <- c(
  yellow_reconstructed_area2x_nonnegative_tracks = file.path(
    ANALYSIS_DIR,
    "data/ou_tracking_fits_yellow_reconstructed_area2x_nonnegative_trackids_min3.csv"
  )
)
DIAGNOSTICS_RDS <- file.path(
  ANALYSIS_DIR,
  "data/ou_report_diagnostics_yellow_reconstructed_area2x_nonnegative_trackids_min3.rds"
)

order_dose_label <- function(label, dose) {
  dose_levels <- sort(unique(dose[is.finite(dose)]))
  factor(label, levels = paste0(dose_levels, " nM"))
}

format_ploidy <- function(x) {
  factor(x, levels = sort(unique(x)))
}

read_ou_fit_dataset <- function(path, dataset) {
  if (!file.exists(path)) {
    return(NULL)
  }
  read_csv(path, show_col_types = FALSE) |>
    mutate(dataset = dataset, source_csv = path)
}

ou_transition <- function(dt, tau, velocity_scale) {
  a <- dt / tau
  phi <- exp(-a)
  one_minus_phi <- -expm1(-a)
  one_minus_phi2 <- -expm1(-2 * a)
  f12 <- tau * one_minus_phi
  q22 <- velocity_scale^2 * one_minus_phi2
  q12 <- velocity_scale^2 * tau * one_minus_phi^2
  q11_unit <- if (a < 1e-4) {
    a^3 / 3 - a^4 / 4 + 7 * a^5 / 60
  } else {
    a - 2 * one_minus_phi + 0.5 * one_minus_phi2
  }
  q11 <- 2 * velocity_scale^2 * tau^2 * q11_unit

  q11 <- max(q11, 0)
  q22 <- max(q22, 0)
  if (q22 > 0 && q11 * q22 < q12^2) {
    q11 <- q12^2 / q22 + .Machine$double.eps
  }

  list(
    F = matrix(c(1, 0, f12, phi), nrow = 2),
    Q = matrix(c(q11, q12, q12, q22), nrow = 2)
  )
}

simulate_ou_track <- function(tau, velocity_scale, n_steps, dt = 1) {
  transition <- ou_transition(dt = dt, tau = tau, velocity_scale = velocity_scale)
  noise_chol <- chol(transition$Q)
  state_x <- c(0, rnorm(1, sd = velocity_scale))
  state_y <- c(0, rnorm(1, sd = velocity_scale))
  rows <- vector("list", n_steps + 1L)
  rows[[1L]] <- tibble(track_step = 0L, x = state_x[[1]], y = state_y[[1]])

  for (step in seq_len(n_steps)) {
    state_x <- drop(transition$F %*% state_x + drop(t(noise_chol) %*% rnorm(2)))
    state_y <- drop(transition$F %*% state_y + drop(t(noise_chol) %*% rnorm(2)))
    rows[[step + 1L]] <- tibble(track_step = step, x = state_x[[1]], y = state_y[[1]])
  }

  bind_rows(rows)
}

simulate_ou_condition <- function(tau, velocity_scale, tau_level, velocity_level, n_tracks = 10L, n_segments = 50L, dt = 1) {
  bind_rows(lapply(seq_len(n_tracks), function(track_index) {
    simulate_ou_track(tau, velocity_scale, n_steps = n_segments, dt = dt) |>
      mutate(track_index = track_index)
  })) |>
    mutate(
      tau = tau,
      velocity_scale = velocity_scale,
      tau_level = tau_level,
      velocity_level = velocity_level,
      condition = paste(tau_level, "tau,", velocity_level, "velocity scale")
    )
}
```

```{r load-data}
fit_inputs <- YELLOW_OU_FITS
ou_fits_raw <- bind_rows(lapply(names(fit_inputs), function(dataset) {
  read_ou_fit_dataset(fit_inputs[[dataset]], dataset)
}))

for (missing_col in setdiff(
  c(
    "n_starts", "seed", "start_id", "best_start_id", "start_tau",
    "start_velocity_scale", "start_obs_noise", "delta_loglik_next_best",
    "n_converged_starts"
  ),
  names(ou_fits_raw)
)) {
  ou_fits_raw[[missing_col]] <- NA
}

ou_fits <- ou_fits_raw |>
  mutate(
    dataset = factor(dataset, levels = names(fit_inputs)),
    ploidy = format_ploidy(ploidy),
    gemcitabine_label = paste0(Gemcitabine, " nM"),
    gemcitabine_label = order_dose_label(gemcitabine_label, Gemcitabine),
    clean_fit_status = if_else(grepl("^converged", fit_status), "converged", fit_status),
    log_likelihood_per_segment = log_likelihood / n_segments,
    log_likelihood_per_track_time = log_likelihood / total_track_time
  ) |>
  arrange(job_id)

best_fit_summary <- ou_fits |>
  group_by(dataset, job_id) |>
  summarize(
    best_log_likelihood = max(log_likelihood, na.rm = TRUE),
    second_best_log_likelihood = suppressWarnings(sort(log_likelihood, decreasing = TRUE, na.last = NA)[2]),
    n_converged_starts = sum(clean_fit_status == "converged", na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    second_best_log_likelihood = if_else(is.finite(second_best_log_likelihood), second_best_log_likelihood, NA_real_),
    delta_loglik_next_best = best_log_likelihood - second_best_log_likelihood
  )

best_fits <- ou_fits |>
  group_by(dataset, job_id) |>
  slice_max(order_by = log_likelihood, n = 1, with_ties = FALSE) |>
  ungroup() |>
  left_join(best_fit_summary, by = c("dataset", "job_id")) |>
  mutate(
    best_start_id = start_id,
    delta_loglik_next_best = coalesce(delta_loglik_next_best.y, delta_loglik_next_best.x),
    n_converged_starts = coalesce(n_converged_starts.y, n_converged_starts.x)
  ) |>
  select(-ends_with(".x"), -ends_with(".y"))

parameter_long <- best_fits |>
  select(
    dataset, job_id, fit_label, ploidy, Gemcitabine, gemcitabine_label, fit_status,
    tau, velocity_scale, effective_diffusivity, obs_noise
  ) |>
  pivot_longer(
    cols = c(tau, velocity_scale, effective_diffusivity, obs_noise),
    names_to = "parameter",
    values_to = "value"
  ) |>
  mutate(
    parameter = factor(
      parameter,
      levels = c("tau", "velocity_scale", "effective_diffusivity", "obs_noise"),
      labels = c("Tau", "Velocity scale", "Effective diffusivity", "Observation noise")
    )
  )

if (!file.exists(DIAGNOSTICS_RDS)) {
  stop(
    "Missing precomputed diagnostics: ", DIAGNOSTICS_RDS, "\n",
    "Run scripts/agentRrunner.sh analyses/K00_GemcitabineExposure_033023/workflow/precompute_ou_report_diagnostics.R",
    call. = FALSE
  )
}
diagnostics <- readRDS(DIAGNOSTICS_RDS)

centered_track_segments <- diagnostics$centered_track_segments
observed_sampled_tracks <- diagnostics$observed_sampled_tracks
simulated_sampled_tracks <- diagnostics$simulated_sampled_tracks
msd_by_step <- diagnostics$msd_by_step
ou_msd_curve <- diagnostics$ou_msd_curve
windowed_msd_by_lag <- diagnostics$windowed_msd_by_lag
step_distribution <- diagnostics$step_distribution
cos_autocorrelation_by_lag <- diagnostics$cos_autocorrelation_by_lag
windowed_cos_autocorrelation_by_lag <- diagnostics$windowed_cos_autocorrelation_by_lag
conditional_endpoints <- diagnostics$conditional_endpoints
track_length_distribution <- diagnostics$track_length_distribution
threshold_sensitivity <- diagnostics$threshold_sensitivity

set.seed(20260508)
illustrative_parameter_levels <- best_fits |>
  summarize(
    low_tau = quantile(.data$tau, 0.10, na.rm = TRUE),
    high_tau = quantile(.data$tau, 0.90, na.rm = TRUE),
    low_velocity_scale = quantile(.data$velocity_scale, 0.10, na.rm = TRUE),
    high_velocity_scale = quantile(.data$velocity_scale, 0.90, na.rm = TRUE)
  )

illustrative_conditions <- tidyr::crossing(
  tau_level = c("Low", "High"),
  velocity_level = c("Low", "High")
) |>
  mutate(
    tau = if_else(.data$tau_level == "Low", illustrative_parameter_levels$low_tau[[1]], illustrative_parameter_levels$high_tau[[1]]),
    velocity_scale = if_else(
      .data$velocity_level == "Low",
      illustrative_parameter_levels$low_velocity_scale[[1]],
      illustrative_parameter_levels$high_velocity_scale[[1]]
    )
  )

illustrative_tracks <- bind_rows(lapply(seq_len(nrow(illustrative_conditions)), function(i) {
  simulate_ou_condition(
    tau = illustrative_conditions$tau[[i]],
    velocity_scale = illustrative_conditions$velocity_scale[[i]],
    tau_level = illustrative_conditions$tau_level[[i]],
    velocity_level = illustrative_conditions$velocity_level[[i]]
  )
})) |>
  mutate(
    condition = factor(
      .data$condition,
      levels = c(
        "Low tau, Low velocity scale", "Low tau, High velocity scale",
        "High tau, Low velocity scale", "High tau, High velocity scale"
      )
    )
  )
```

## Inputs

This report reads:

```text
`r paste(fit_inputs[file.exists(fit_inputs)], collapse = "\n")`
`r DIAGNOSTICS_RDS`
```

These are condition-level OU velocity-model fits and precomputed diagnostics for
`yellow_reconstructed_area2x_nonnegative_tracks`. Multistart outputs have one
row per start attempt, so the main parameter plots use the highest-likelihood
start for each condition.

```{r fit-status-table}
ou_fits |>
  count(dataset, clean_fit_status, name = "n_start_attempts")
```

## Generative Model Illustration

This simulation combines low/high tau with low/high velocity scale using the
same OU position/velocity transition form as the fitted model. Each condition
shows 10 simulated tracks with 50 segments.

```{r generative-model-track-plots, fig.width=11, fig.height=9}
ggplot(
  illustrative_tracks,
  aes(
    x = .data$x,
    y = .data$y,
    group = .data$track_index
  )
) +
  geom_path(linewidth = 0.45, alpha = 0.65, color = "grey25") +
  geom_point(aes(color = .data$track_step), size = 0.35, alpha = 0.75) +
  scale_color_viridis_c() +
  coord_equal() +
  facet_wrap(~ condition, ncol = 2) +
  labs(
    x = "X position",
    y = "Y position",
    color = "Step"
  ) +
  theme_bw()
```

## Fitted Parameters

```{r fitted-parameters-by-dose, fig.width=11, fig.height=8}
ggplot(
  parameter_long |> filter(is.finite(value)),
  aes(x = gemcitabine_label, y = value, color = ploidy, group = interaction(dataset, ploidy), linetype = dataset)
) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ parameter, scales = "free_y", ncol = 2) +
  labs(
    x = "Gemcitabine",
    y = "Fitted value",
    color = "Ploidy",
    linetype = "Dataset"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Fit Support

```{r multistart-diagnostics, fig.width=10, fig.height=4}
multistart_data <- ou_fits |>
  count(dataset, job_id, ploidy, Gemcitabine, gemcitabine_label, name = "observed_starts") |>
  left_join(
    best_fits |>
      select(dataset, job_id, best_start_id, n_starts, delta_loglik_next_best, n_converged_starts),
    by = c("dataset", "job_id")
  ) |>
  filter(is.finite(delta_loglik_next_best))

if (nrow(multistart_data) > 0) {
  ggplot(multistart_data, aes(x = gemcitabine_label, y = delta_loglik_next_best, color = ploidy, group = interaction(dataset, ploidy), linetype = dataset)) +
    geom_line(linewidth = 0.7, alpha = 0.8) +
    geom_point(size = 2) +
    labs(
      x = "Gemcitabine",
      y = "Best log likelihood - next best",
      color = "Ploidy",
      linetype = "Dataset"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

```{r fit-support, fig.width=11, fig.height=5}
support_long <- best_fits |>
  select(dataset, job_id, ploidy, Gemcitabine, gemcitabine_label, n_segments, total_track_time) |>
  pivot_longer(
    cols = c(n_segments, total_track_time),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = factor(
      metric,
      levels = c("n_segments", "total_track_time"),
      labels = c("OU segments", "Total track time")
    )
  )

ggplot(support_long, aes(x = gemcitabine_label, y = value, color = ploidy, group = interaction(dataset, ploidy), linetype = dataset)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  labs(
    x = "Gemcitabine",
    y = "Support",
    color = "Ploidy",
    linetype = "Dataset"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r normalized-log-likelihood, fig.width=11, fig.height=5}
likelihood_long <- best_fits |>
  select(
    dataset, job_id, ploidy, Gemcitabine, gemcitabine_label,
    log_likelihood_per_segment, log_likelihood_per_track_time
  ) |>
  pivot_longer(
    cols = c(log_likelihood_per_segment, log_likelihood_per_track_time),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = factor(
      metric,
      levels = c("log_likelihood_per_segment", "log_likelihood_per_track_time"),
      labels = c("Log likelihood / OU segment", "Log likelihood / total track time")
    )
  )

ggplot(
  likelihood_long |> filter(is.finite(value)),
  aes(x = gemcitabine_label, y = value, color = ploidy, group = interaction(dataset, ploidy), linetype = dataset)
) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  labs(
    x = "Gemcitabine",
    y = "Normalized log likelihood",
    color = "Ploidy",
    linetype = "Dataset"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Observed vs Simulated Tracks

For the selected doses below, simulated tracks use the best OU fit for the
matching ploidy and dose and preserve the sampled observed track-length
distribution.

```{r observed-vs-simulated-centered-tracks, fig.width=14, fig.height=10}
track_overlay_compare <- bind_rows(observed_sampled_tracks, simulated_sampled_tracks) |>
  mutate(source = factor(.data$source, levels = c("Observed", "OU simulated")))

ggplot(
  track_overlay_compare,
  aes(
    x = .data$centered_x,
    y = .data$centered_y,
    group = interaction(.data$source, .data$site_id, .data$migration_track_id)
  )
) +
  geom_path(linewidth = 0.16, alpha = 0.18, color = "grey25") +
  coord_equal() +
  facet_grid(ploidy ~ gemcitabine_label + source) +
  labs(
    x = "X displacement from segment start (px)",
    y = "Y displacement from segment start (px)"
  ) +
  theme_bw() +
  theme(strip.text.x = element_text(size = 8))
```

## Mean Squared Displacement

Mean squared displacement is computed from each track's first observed position
to the nth observed step of the same track. Dashed lines show the fitted OU
expectation, including fitted observation noise.

```{r mean-squared-displacement, fig.width=11, fig.height=6}
ggplot() +
  geom_line(
    data = msd_by_step |> filter(.data$source == "Observed"),
    aes(
      x = .data$track_step,
      y = .data$mean_squared_displacement_px2,
      color = .data$ploidy,
      group = .data$ploidy
    ),
    linewidth = 0.7,
    alpha = 0.9
  ) +
  geom_point(
    data = msd_by_step |> filter(.data$source == "Observed"),
    aes(
      x = .data$track_step,
      y = .data$mean_squared_displacement_px2,
      color = .data$ploidy,
      size = .data$n_tracks
    ),
    alpha = 0.65
  ) +
  geom_line(
    data = ou_msd_curve,
    aes(
      x = .data$track_step,
      y = .data$mean_squared_displacement_px2,
      color = .data$ploidy,
      group = .data$ploidy
    ),
    linewidth = 0.65,
    alpha = 0.75,
    linetype = "dashed"
  ) +
  facet_wrap(~ gemcitabine_label, scales = "free_y", ncol = 5) +
  scale_size_continuous(trans = "log10", range = c(0.4, 2.4), breaks = c(10, 100, 1000, 10000)) +
  labs(
    x = "Step from initial tracked position",
    y = "Mean squared displacement (px^2)",
    color = "Ploidy",
    size = "Tracks",
    caption = "Solid = observed; dashed = fitted OU expectation."
  ) +
  theme_bw()
```

```{r time-stratified-msd, fig.width=12, fig.height=8}
ggplot(
  windowed_msd_by_lag,
  aes(
    x = .data$lag_frames,
    y = .data$mean_squared_displacement_px2,
    color = .data$ploidy,
    linetype = .data$source,
    group = interaction(.data$ploidy, .data$source)
  )
) +
  geom_line(linewidth = 0.65, alpha = 0.9) +
  geom_point(aes(size = .data$n_displacements), alpha = 0.55) +
  facet_grid(frame_window ~ gemcitabine_label, scales = "free_y") +
  scale_size_continuous(trans = "log10", range = c(0.3, 2.0), breaks = c(10, 100, 1000, 10000)) +
  labs(
    x = "Lag (frames)",
    y = "Mean squared displacement (px^2)",
    color = "Ploidy",
    linetype = "Source",
    size = "Displacements"
  ) +
  theme_bw()
```

## Step-Length Distribution

One-frame step lengths are capped at 50 px for display, so the rightmost bin
includes all steps with displacement `>=50` px.

```{r observed-vs-simulated-step-lengths, fig.width=12, fig.height=7}
ggplot(
  step_distribution,
  aes(x = .data$capped_step_distance_px, fill = .data$source)
) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 2, position = "identity", alpha = 0.35, boundary = 0) +
  facet_grid(ploidy ~ gemcitabine_label) +
  scale_x_continuous(limits = c(0, 50), breaks = seq(0, 50, 10)) +
  labs(
    x = "One-frame step length (px, capped at 50)",
    y = "Density",
    fill = "Source"
  ) +
  theme_bw()
```

```{r time-stratified-step-lengths, fig.width=12, fig.height=8}
ggplot(
  step_distribution |> filter(!is.na(.data$frame_window)),
  aes(x = .data$capped_step_distance_px, fill = .data$source)
) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 2, position = "identity", alpha = 0.35, boundary = 0) +
  facet_grid(frame_window ~ ploidy) +
  scale_x_continuous(limits = c(0, 50), breaks = seq(0, 50, 10)) +
  labs(
    x = "One-frame step length (px, capped at 50)",
    y = "Density",
    fill = "Source"
  ) +
  theme_bw()
```

## Directional Persistence

Directional autocorrelation compares each step vector with later step vectors:

\[
C(k)=E\left[\frac{\Delta x_t \cdot \Delta x_{t+k}}{\|\Delta x_t\|\|\Delta x_{t+k}\|}\right].
\]

If the fitted OU model captures directional persistence, the simulated curves
should follow the observed decay. Slower decay in 4N would support the
interpretation that fitted 4N tau reflects persistent motion rather than only
larger one-frame steps.

```{r directional-persistence-cosine, fig.width=12, fig.height=7}
ggplot(
  cos_autocorrelation_by_lag,
  aes(
    x = .data$lag_frames,
    y = .data$mean_velocity_cos_autocorrelation,
    color = .data$ploidy,
    linetype = .data$source,
    group = interaction(.data$ploidy, .data$source)
  )
) +
  geom_hline(yintercept = 0, linewidth = 0.25, color = "grey60") +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  geom_point(aes(size = .data$n_pairs), alpha = 0.65) +
  facet_wrap(~ gemcitabine_label, ncol = 5) +
  scale_size_continuous(trans = "log10", range = c(0.4, 2.4), breaks = c(10, 100, 1000, 10000)) +
  labs(
    x = "Step-vector lag (frames)",
    y = "Mean cosine autocorrelation",
    color = "Ploidy",
    linetype = "Source",
    size = "Pairs"
  ) +
  theme_bw()
```

```{r time-stratified-directional-persistence, fig.width=12, fig.height=8}
ggplot(
  windowed_cos_autocorrelation_by_lag,
  aes(
    x = .data$lag_frames,
    y = .data$mean_velocity_cos_autocorrelation,
    color = .data$ploidy,
    linetype = .data$source,
    group = interaction(.data$ploidy, .data$source)
  )
) +
  geom_hline(yintercept = 0, linewidth = 0.25, color = "grey60") +
  geom_line(linewidth = 0.65, alpha = 0.9) +
  geom_point(aes(size = .data$n_pairs), alpha = 0.55) +
  facet_grid(frame_window ~ gemcitabine_label) +
  scale_size_continuous(trans = "log10", range = c(0.3, 2.0), breaks = c(10, 100, 1000, 10000)) +
  labs(
    x = "Step-vector lag (frames)",
    y = "Mean cosine autocorrelation",
    color = "Ploidy",
    linetype = "Source",
    size = "Pairs"
  ) +
  theme_bw()
```

## Conditional Turning

Current displacement vectors are rotated to point right, then the next
one-frame displacement endpoint is plotted in that aligned coordinate system.
This directly checks whether the OU simulation reproduces observed turning
behavior conditional on the current step size.

```{r conditional-next-step-endpoints, fig.width=13, fig.height=9}
ggplot(
  conditional_endpoints |> filter(.data$current_step_length_bin %in% c("5-10 px", "10-20 px")),
  aes(x = .data$next_dx_aligned, y = .data$next_dy_aligned, color = .data$source)
) +
  geom_hline(yintercept = 0, linewidth = 0.2, color = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.2, color = "grey70") +
  stat_density_2d(linewidth = 0.35, alpha = 0.8, contour_var = "ndensity") +
  coord_equal(xlim = c(-40, 60), ylim = c(-50, 50)) +
  facet_grid(ploidy + current_step_length_bin ~ gemcitabine_label) +
  labs(
    x = "Next dx after aligning current step to +x (px)",
    y = "Next dy after aligning current step to +x (px)",
    color = "Source"
  ) +
  theme_bw()
```

## Track-Length and Threshold QC

```{r track-length-distribution, fig.width=11, fig.height=6}
ggplot(track_length_distribution, aes(x = .data$gemcitabine_label, y = .data$n_tracks, fill = .data$track_length_bin)) +
  geom_col(position = "fill", width = 0.75) +
  facet_wrap(~ ploidy, ncol = 1) +
  labs(
    x = "Gemcitabine",
    y = "Fraction of tracks",
    fill = "Track frames"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r min-track-length-sensitivity, fig.width=11, fig.height=7}
threshold_plot <- threshold_sensitivity |>
  select(
    min_track_frames, ploidy, Gemcitabine, gemcitabine_label,
    n_tracks, median_step_distance_px, msd_step_3, msd_step_5, msd_step_10
  ) |>
  pivot_longer(
    cols = c(n_tracks, median_step_distance_px, msd_step_3, msd_step_5, msd_step_10),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    min_track_frames = factor(paste0(">=", .data$min_track_frames)),
    metric = factor(
      .data$metric,
      levels = c("n_tracks", "median_step_distance_px", "msd_step_3", "msd_step_5", "msd_step_10"),
      labels = c("Tracks retained", "Median step length", "MSD step 3", "MSD step 5", "MSD step 10")
    )
  )

ggplot(
  threshold_plot,
  aes(x = .data$gemcitabine_label, y = .data$value, color = .data$ploidy, group = interaction(.data$ploidy, .data$min_track_frames))
) +
  geom_line(aes(linetype = .data$min_track_frames), linewidth = 0.65, alpha = 0.8) +
  geom_point(aes(shape = .data$min_track_frames), size = 1.6, alpha = 0.85) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  labs(
    x = "Gemcitabine",
    y = "Value",
    color = "Ploidy",
    linetype = "Minimum frames",
    shape = "Minimum frames"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The strongest empirical support is for a dispersal difference: step lengths and
MSD-like summaries show how far cells move under the staged yellow-track input.
The OU tau parameter is a model-based persistence estimate and is more
vulnerable to short tracks, track-length filtering, and unmodeled
time-varying biology. Current results should therefore be phrased as greater
4N dispersal with OU fits suggesting increased persistence, not as a standalone
proof that tau is stable across QC choices.

## Deferred Confounders

Alive/dead classification and proliferation adjustment are not included in this
report because no staged joinable inputs are present here. Expected future
inputs are:

- a per-cell or per-track alive/dead table joinable by well, position, frame,
  object or track ID;
- a per-well proliferation summary joinable by well and condition.

Once staged, the next checks are live-only migration summaries, alive/dead
composition by ploidy and dose, and migration estimates stratified or adjusted
by proliferation.

## Fit Table

```{r fit-table}
best_fits |>
  select(
    dataset, job_id, fit_label, ploidy, Gemcitabine, fit_status,
    tau, velocity_scale, effective_diffusivity, obs_noise,
    log_likelihood, log_likelihood_per_segment, log_likelihood_per_track_time,
    n_starts, seed, best_start_id, start_tau, start_velocity_scale,
    start_obs_noise, delta_loglik_next_best, n_converged_starts,
    n_segments, total_track_time
  )
```
